A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)

### 2019-11-12 Simulation/Writeup/preprint.Rnw
Decided to split the Research Report and the final article (Thesis). RR is in Writeup folder of Simulation now.

%% recommended packages
\usepackage{thumbpdf,lmodern}

@@

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
\Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
\Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}

%% - \Abstract{} almost as usual
\Abstract{
  This Research Report contains the Introduction and Methods section of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. ("There is no page limit, nor a limit on the number of figures or tables", see \url{https://www.jstatsoft.org/pages/view/authors}.) I have chosen to use the second format from the Research Report guidelines: "It is the first half of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and or statistical procedures". The goal was to develop novel methodology and guidelines for evaluating multiple imputation methods, and implement these in an interactive evaluation framework for multiple imputation: \pkg{ShinyMICE}.
  }

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{multiple imputation, evaluation methodology, \pkg{ShinyMICE}, \pkg{mice}, \proglang{R}}
\Plainkeywords{multiple imputation, evaluation methodology, ShinyMICE, mice, R}

@@

%% - In principle "as usual".
%% - But should typically have some discussion of both _software_ and _methods_.
%% - Use \proglang{}, \pkg{}, \fct{}, and \code{} markup throughout the manuscript.
%% - If such markup is in (sub)section titles, a plain text version has to be
%%   added as well.
%% - All software mentioned should be properly \cite-d.
%% - All abbreviations should be introduced.
%% - Unless the expansions of abbreviations are proper names (like "Journal
%%   of Statistical Software" above) they should be in sentence case (like
%%   "generalized linear models" below).

\section[Introduction: Multiple Imputation Methodology]{Introduction: Multiple Imputation Methodology} \label{sec:intro} % curly brackets can handle typesetting

% \begin{leftbar}
% The introduction is in principle ``as usual''. However, it should usually embed
% both the implemented \emph{methods} and the \emph{software} into the respective
% relevant literature. For the latter both competing and complementary software
% should be discussed (within the same software environment and beyond), bringing
% out relative (dis)advantages. All software mentioned should be properly
% \verb|\cite{}|d. (See also Appendix~\ref{app:bibtex} for more details on
% \textsc{Bib}{\TeX}.) Example: % \citep[][Chapter~7.4]{Venables+Ripley:2002} 
% 
% For writing about software JSS requires authors to use the markup
% \verb|\proglang{}| (programming languages and large programmable systems),
% \verb|\pkg{}| (software packages), \verb|\code{}| (functions, commands,
% arguments, etc.). If there is such markup in (sub)section titles (as above), a
% plain text version has to be provided in the {\LaTeX} command as well. Below we
% also illustrate how abbrevations should be introduced and citation commands can
% be employed. See the {\LaTeX} code for more details.
% \end{leftbar}

@@

With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' \emph{data generating model} \citep{buur18}. Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid. Such evaluation measures are currently missing or under-developed in MI software, like the world leading \proglang{R} package \pkg{mice} \citep{mice}. Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.

@@

The aim of this paper is to provide applied researchers an introduction to the interactive evaluation device \pkg{ShinyMICE}. The intended audience consists of empirical researchers and statisticians who (want to) use multiple imputation to solve missing data problems. Basic familiarity with multiple imputation methodology is assumed. For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}. For the theoretical foundation of MI, see \cite{rubin87}. All programming code used in this paper is available in the file 'XYZ.R' along with the manuscript, and on Github repository 'XYZ'.

@@

The \proglang{R} \pkg{Shiny} application introduced in this paper is developed with the aim to integrate it into the \pkg{mice} environment. \pkg{ShinyMICE} therefore follows notation and conventions of \cite{mice}. For an overview of the deviations from the 'original' notation by \cite{rubin87}, see \cite{buur18}.

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values is denoted as $Y_{obs}$; the missing part of $Y$ is refered to as $Y_{mis}$. Response indicator $R$ shows whether a data value in $Y$ is missing or observed. The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.

% "Let Y denote the n x p matrix containing the data values on p variables for all n units in the sample. We define the response indicator R as an n x p 0-1 matrix. The elements of Y and R are denoted by y i j and r i j , respectively, where i = 1 , ... , n and j = 1 , ... , p . If y i j is observed, then r i j = 1 , and if y i j is missing, then r i j = 0 ." ... The observed data are collectively denoted by Y o b s . The missing data are collectively denoted as Y m i s , and contain all elements y i j where r i j = 0 . When taken together Y = ( Y o b s , Y m i s ) contain the hypothetically complete data." \cite[par.~2.2]{buur18}

% The data are said to be MCAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | $\psi$ ) (2.1) so the probability of being missing depends only on some parameters $\psi$ , the overall probability of being missing. The data are said to be MAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | Y o b s , $\psi$ ) (2.2) so the missingness probability may depend on observed information, including any design factors. Finally, the data are MNAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) (2.3) does not simplify, so here the probability to be missing also depends on unobserved information, including Y m i s itself. \cite[par.~2.2]{buur18}

Terminology (MCAR, MAR, MNAR).

Blue points are observed, the red points are imputed.

@@

- missingness mechanims, ignorability

"The practical importance of the distinction between MCAR, MAR and MNAR is that it clarifies the conditions under which we can accurately estimate the scientifically interesting parameters without the need to know $\psi$" \cite[par.~2.2]{buur18}.

- Rubin's rules

- FCS vs. JM?

\subsection{Theoretical Background from Thesis Proposal}

The validity of the MI solution depends on numerous assumptions that cannot be verified from the observed data alone. So instead of statistical tests for assumptions, evaluation procedures have been developed. For the following assumptions, no reliable procedure has been proposed and/or implemented: 1) \emph{ignorability} of the \emph{missingness mechanism} \citep{rubin87}; 2) \emph{congeneality} of the imputation models \citep{meng94}; and 3) \emph{compatibility} of the MI modeling procedure \citep{rubin96}.

\begin{enumerate}
\item A missingness mechanism is said to be ignorable when the probability to be missing does not depend on the missing data itself. Violation of this assumption can gravely affect inferences. Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses. Some practical guidelines exist (e.g., \citep{nguy17}), but current MI software does not facilitate this methodology for empirical researchers.  % use '\citep[see][for further details]{...}' for '(see Author Year for further details)'
%
\item Congeneal imputation models capture all required relations between observed and missing parts of the data. The extent to which this has been successful can be evaluated by plotting conditional distributions \citep{abay08}. Such visualizations are available in MICE, but subsequent statistical tests to quantify the relations with covariates are not provided. 
% Additionally, there is potential to assess model fit by means of \emph{over-imputation} \citep{buur18} or *double robustness* \citep{bang05}-- topics that are only persued if time permits.
%
\item The third assumption is met when the MI algorithm converges to a stable distribution. However, conventional measures to diagnose convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R}$ --are not applicable on multiply imputed data \citep{lace07}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11}. Visually assessing convergence is not only difficult to the untrained eye, it might also be futile. The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}. Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
\end{enumerate}

@@

\subsection{Approach from Thesis Proposal}

Initially, the research project will consist of an investigation into algorithmic convergence of MI algorithms. I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

Subsequently, I will use\proglang{R}Shiny \citep{shiny} to implement the convergence indicator and existing evaluation measures in \pkg{ShinyMICE}, see Figure 1. The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$~tests or $t$~tests).

A working beta version of \pkg{ShinyMICE} will be considered a sufficient milestone to proceed with writing a technical paper on the methodology and the software. I will submit the paper for publication in \emph{Journal of Statistical Software}. Finally, \pkg{ShinyMICE} will be integrated into the existing MICE environment, and a vignette for applied researchers will be written.

The \proglang{R} code and documentation of this project will be open source (available on Github). Since the study does not require the use of unpublished empirical data, I expect that the FETC will grant the label exempt.

% \includegraphics[height=80mm,keepaspectratio]{impression} % this is not correct yet, it needs a figure label and stuff

%% -- Methods: Sensitivity Analyses ---------------------------------------------------------------
\subsection{Sensitivity Analyses} \label{sec:sensitivity}

% In general: "The MAR assumption can never be tested from the observed data. One can however check whether the imputations created by MICE algorithm are plausible" \cite[p.~42]{mice}.

- Current guideine: "Plot densities of both the observed and imputed values of all variables to see whether the imputations are reasonable. Differences in the densities between the observed and imputed values may suggest a problem that needs to be further checked" \cite[p.~43]{mice}. 
% FIMD on sensitivity analyses, see https://stefvanbuuren.name/fimd/sec-sensitivity.html

- Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses. See \citep{nguy17} for guidelines.

- Forked MICE with MNAR sensitivity analyses
% see \url{https://thestatsgeek.com/2018/06/07/missing-not-at-random-sensitivity-analysis-with-fcs-multiple-imputation/}.

- Nothing within the scope of this thesis.

%% -- Methods: Visualizations ---------------------------------------------------------------
\subsection{Data Visualization} \label{sec:visualization}

% In general: "An important step in multiple imputation is to assess whether imputations are plausible. Imputations should be values that could have been obtained had they not been missing. Imputations should be close to the data. Data values that are clearly impossible (e.g., negative counts, pregnant fathers) should not occur in the imputed data. Imputations should respect relations between variables, and reflect the appropriate amount of uncertainty about their 'true' values. Diagnostic checks on the imputed data provide a way to check the plausibility of the imputations" \cite[p.~11]{mice}.

- Existing plot functions to evaluate relations between observed and missing parts of the data (\pkg{mice} functions \fct{hist}, \fct{xyplot}, and \fct{stripplot}). 
% "The mice package contains several graphic functions that can be used to gain insight into the correspondence of the observed and imputed data: \fct{bwplot}, \fct{stripplot}, \fct{densityplot} and \fct{xyplot}." \cite[par.~6.6]{buur18}

% "It is often useful to inspect the distributions of original and the imputed data. One way of doing this is to use the function \fct{stripplot} \dots  The differences between the red points represents our uncertainty about the true (but unknown) values. \dots Under MCAR, univariate distributions of the observed and imputed data are expected to be identical. Under MAR, they can be different, both in location and spread, but their multivariate distribution is assumed to be identical." \cite[p.~12]{mice}.

- Improve plots using \pkg{lattice} package functionalities.

- Subsequent statistical tests to quantify the relations with covariates.
% "Bondarenko and Raghunathan (2016) [see https://www.ncbi.nlm.nih.gov/pubmed/26952693] proposed a more refined diagnostic tool that aims to compare the distributions of observed and imputed data conditional on the missingness probability. The idea is that under MAR the conditional distributions should be similar if the assumed model for creating multiple imputations has a good fit. An example is created as ..."  \cite[par.~6.6]{buur18}
% 
% fit <- with(imp, glm(ici(imp) ~ age + bmi + hyp + chl,
%                      family = binomial))
% ps <- rep(rowMeans(sapply(fit$analyses, fitted.values)),
%           imp$m + 1)
% xyplot(imp, bmi ~ ps | as.factor(.imp),
%        xlab = "Probability that record is incomplete",
%        ylab = "BMI", pch = c(1, 19), col = mdc(1:2))

- Use evaluation of MI methods in SAS as example
% see https://onlinelibrary.wiley.com/doi/full/10.1111/1467-9574.00219

- Nothing within the scope of this thesis.

%% -- Methods: Convergence ---------------------------------------------------------------
\subsection{Convergence Diagnostic} \label{sec:convergence}

@@

- Traceplots

@@

- $\widehat{R}$, but too stringent (new) threshold, and assumption of overdispersed initial values of imputation chains not met.
% Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see \citep{veht19}.

- Autocorrelation. Schafer (1997, p. 129) wrote on worst linear statistic. We could calculate the autocorrelation of that statistic to know that the algorithm converged elsewhere too. See autocorr function plot in sas of worst linear function.
% Worst linear function in SAS see \url{https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mi_sect027.htm}.
% Note: we're talking about missing data only, not the combined data (that autocorrelation is very high, as is the autocorrelation of deductively imputed values, like in the texp example). 
% "Applications of MICE with lowly correlated data therefore inject a lot of noise into the system. Hence, the auto-correlation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent. Thus, the incorporation of noise into the imputed data has pleasant side-effect of speeding up convergence" \citep{buur18}, par. 4.5.

@@

- $\hat{R}$ threshold: Replicate simulation study and build a decision rule to solve the problem with $\hat{R}$.
% "The monitoring statistic was computed for mean monthly earnings at each iteration in chains of length k = 200. Since calculation of the statistic requires M parallel sequences, m = 5 such chains were constructed. This value of m was informed by the preferred choice given in the literature on multiple imputation. The monitoring statistic computed at each iteration is presented in Figures 15 to 17 for each of the three missingness mechanisms. The red vertical line denotes ten iterations" \cite[p.~49]{lace07}.

@@

\section{Simulation study}

\subsection{Simulation: Introduction}

\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results. The length of the research report should be maximally 2500 words of text (without references list and or tables and figures). Please do not include appendices, and no more than 6 tables or figures. Table and Figure captions do not count towards the word limit. An abstract may be included, but is not necessary.}

R hat not working on MI data? "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}. Aim to replicate Lacerda et al. that R hat will not be smaller than 1.1 before iteration nr. 50.

As recommended by ... Rhat is computed as .... to be able to detect ... in the tails of the distribution.

R hat is not appropriate because it assumes overdispersed initial values, which means that the initial values of the m imputation chains should be 'far away' from the distribution that the chains are converging to (the modus of the distribution). With MI procedures, initial values are chosen such that ... or maybe even estimated from the observed data. This means that at most one initial value is necessary, but probably none at al. I should look this up.

Convergence can also be interpreted as stable over iterations, or non-recurring. Non-recurrence can be evaluated with autocorrelation. Autocorrelation shows how dependent subsequent draws of an imputation chain are on the previous value. If there is a lot of dependence, draws at e.g. iteration five are significantly correlated with the value of the first draw. Ideally, we want the chains to intermingle nicely, without trends witihin chains. Autocorrelation above the confidence level indicate dependence within chains.

\subsection{Simulation: Methodology}

Explain diagnostics here (see below).

We want to evaluate whether the imputation procedure has converged. We can apply GR statistic Rhat to the first two moments of the variables of interest (mean and variance). The primary research interest is in determining whether Rhat is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data. The simulation diagnostics are as recommended by Stef van Buuren \cite{buur18}. That is, absolute bias confidence interval width, and empirical coverage rate (coverage probability?) across simulations.

Hypothesis based on \cite{lace07} is that the conventional acceptable level of Rhat is too strict for MI data. We expect that the simulation diagnostics will indicate proper imputations before Rhat will.

Evaluate: bias, variance, and cov rate to find problems that you might miss otherwise. There is no tradeoff between them. Also look at distributional char., and plausibility of imp. values, see \cite{vinknd}.

@@

\subsection{Simulation: Simulation Approach}

Set-up: simulate data, impute missing data points with varying number of iterations. Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and the convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate (coverage probability?) across simulations. Convergence is evaluated with $\hat{R}$, or potential scale reduction factor. Proposed by Gelman and Rubin (1992), but since developed into ...
% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p. 4)

In this study, 1000 MCMC simulations are performed.

\begin{Code}
# pseudocode of similation 
simulate data with missingness
for (number of iterations from 1 to 100)
  impute the missingness
  compute convergence diagnostic
  perform analysis
  pool results
  compute simulation diagnostics
save diagnostics
\end{Code}

Simulation code is available in online appendix XYZ on Github.

\subsection{Simulation: Results}

Rhat is not perfect for MI data. The value shoudl theoretically not be smaller than one, yet it happened several times in this study. How could Rhat smaller than 1 occur? The number of simulations is smaller than in 'regular' MCMC processes. Therefore, the '$n-1/n)$' correction factor can influence the estimated potential scale reduction facotr. This downwards bias is in the opposite direction than expected.

\subsection{Simulation Discussion}

"The mixture-of-sequences variance, V, should stabilize as a function of n. (Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)" \cite[p~438]{broo98}.

We would like to have convergence measures for multivariable statistics (scalars?) of interest. This is, however, dependent of the complete data model. The eigenvector decomposition methof proposed by McKay (?) should be implemented. I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this apprach could be implemented.

>>>>

%% recommended packages
\usepackage{thumbpdf,lmodern, amsmath}

@@

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
\Plaintitle{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
\Shorttitle{Convergence Diagnostics for MICE}

%% - \Abstract{} almost as usual
\Abstract{
  This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I have chosen to use the first format from the Research Report guidelines: \emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results. The length of the research report should be maximally 2500 words of text (without references list and or tables and figures). Please do not include appendices, and no more than 6 tables or figures. Table and Figure captions do not count towards the word limit. An abstract may be included, but is not necessary.}
%"It is the first half of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and or statistical procedures". The goal was to develop novel methodology and guidelines for evaluating multiple imputation methods, and implement these in an interactive evaluation framework for multiple imputation: \pkg{ShinyMICE}.
  }

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{multiple imputation, convergence, \pkg{mice}, \proglang{R}}
\Plainkeywords{multiple imputation, convergence, mice, R}

@@

\section{Introduction} \label{sec:intro} % curly brackets can handle typesetting

@@

Missing data is ubiquitous. MICE can solve stuff.

Validity of inference MCMC algorithms is threatened by non-convergence. But it is not known whether conventional convergence diagnostics work for MI data. Aim of this paper is to investigate convergence properties of MI algorithms by means of simulation. Results of simulation study can be used as guideline for applied researchers. %I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

% With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' \emph{data generating model} \citep{buur18}. Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid. Such evaluation measures are currently missing or under-developed in MI software, like the world leading \proglang{R} package \pkg{mice} \citep{mice}. Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.

@@

The aim of this paper is to provide applied researchers a tool to assess convergence of MICE algorithms. The intended audience consists of empirical researchers and statisticians who use multiple imputation to solve missing data problems. Basic familiarity with multiple imputation methodology is assumed. For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}. For the theoretical foundation of MI, see \cite{rubin87}. All programming code used in this paper is available in the file 'XYZ.R' along with the manuscript, and on Github repository 'XYZ'.

@@

The convergence diagnostic introduced in this paper is developed with the aim to integrate it into the \proglang{R} package \pkg{mice} environment. It therefore follows notation and conventions of \cite{mice}. For an overview of the deviations from the 'original' notation by \cite{rubin87}, see \cite{buur18}.

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$. Response indicator $R$ shows whether a data value in $Y$ is missing or observed. The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.

% "Let Y denote the n x p matrix containing the data values on p variables for all n units in the sample. We define the response indicator R as an n x p 0-1 matrix. The elements of Y and R are denoted by y i j and r i j , respectively, where i = 1 , ... , n and j = 1 , ... , p . If y i j is observed, then r i j = 1 , and if y i j is missing, then r i j = 0 ." ... The observed data are collectively denoted by Y o b s . The missing data are collectively denoted as Y m i s , and contain all elements y i j where r i j = 0 . When taken together Y = ( Y o b s , Y m i s ) contain the hypothetically complete data." \cite[par.~2.2]{buur18}
% The data are said to be MCAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | $\psi$ ) (2.1) so the probability of being missing depends only on some parameters $\psi$ , the overall probability of being missing. The data are said to be MAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | Y o b s , $\psi$ ) (2.2) so the missingness probability may depend on observed information, including any design factors. Finally, the data are MNAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) (2.3) does not simplify, so here the probability to be missing also depends on unobserved information, including Y m i s itself. \cite[par.~2.2]{buur18}

% Also introduce: 
% 
% - Terminology (MCAR, MAR, MNAR)?
% 
% - Blue points are observed, the red points are imputed?

@@

Non-convergence has two interpretations [look up source!!!]: mixing between chains and stability over iterations within chains. Diagnose non-convergence: $\widehat{R}$, auto-correlation, MC error, Geweke, the other one. However, existing methodology may not work on MI data. Focus on $\widehat{R}$.

As recommended by \cite{veth19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution.

Why is R hat not applicable on MI data? R hat is not appropriate because it assumes over-dispersed initial values, which means that the initial values of the m imputation chains should be 'far away' from the distribution that the chains are converging to (the mode of the distribution). With MI procedures, initial values are chosen such that ... or maybe even estimated from the observed data. This means that at most one initial value is necessary, but probably none at al. I should look this up. And, "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}. Empirical finding suggests otherwise: that R hat will not be smaller than 1.1 before iteration number. 50. Therefore, the aim is to replicate empirical finding Lacerda et al.

Convergence can also be interpreted as stable over iterations, or non-recurring. Non-recurrence can be evaluated with auto-correlation. Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value. If there is a lot of dependence, draws at e.g. iteration five are significantly correlated with the value of the first draw. Ideally, we want the chains to intermingle nicely, without trends within chains. Auto-correlation above the confidence level indicate dependence within chains.

% Also introduce:
% 
% - missingness mechanims, ignorability?
% % "The practical importance of the distinction between MCAR, MAR and MNAR is that it clarifies the conditions under which we can accurately estimate the scientifically interesting parameters without the need to know $\psi$" \cite[par.~2.2]{buur18}.
% 
% - Rubin's rules?
% 
% - FCS vs. JM?
%

@@

The validity of the MI solution depends on numerous assumptions that cannot be verified from the observed data alone. So instead of statistical tests for assumptions, evaluation procedures have been developed. For the following assumptions, no reliable procedure has been proposed and/or implemented: 1) \emph{ignorability} of the \emph{missingness mechanism} \citep{rubin87}; 2) \emph{congeniality} of the imputation models \citep{meng94}; and 3) \emph{compatibility} of the MI modeling procedure \citep{rubin96}.

% \begin{enumerate}
% \item A missingness mechanism is said to be ignorable when the probability to be missing does not depend on the missing data itself. Violation of this assumption can gravely affect inferences. Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses. Some practical guidelines exist (e.g., \citep{nguy17}), but current MI software does not facilitate this methodology for empirical researchers.  % use '\citep[see][for further details]{...}' for '(see Author Year for further details)'
% %
% \item Congeneal imputation models capture all required relations between observed and missing parts of the data. The extent to which this has been successful can be evaluated by plotting conditional distributions \citep{abay08}. Such visualizations are available in MICE, but subsequent statistical tests to quantify the relations with covariates are not provided. 
% % Additionally, there is potential to assess model fit by means of \emph{over-imputation} \citep{buur18} or *double robustness* \citep{bang05}-- topics that are only persued if time permits.
% %
%\item 
The third assumption is met when the MI algorithm converges to a stable distribution. However, conventional measures to diagnose convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R}$ --are not applicable on multiply imputed data \citep{lace07}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11}. Visually assessing convergence is not only difficult to the untrained eye, it might also be futile. The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}. Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
%\end{enumerate}

@@

- Trace-plots

@@

- $\widehat{R}$, but too stringent (new) threshold, and assumption of over-dispersed initial values of imputation chains not met.
% Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see \citep{veht19}.

- Auto-correlation. Schafer (1997, p. 129) wrote on worst linear statistic. We could calculate the auto-correlation of that statistic to know that the algorithm converged elsewhere too. See autocorr function plot in SAS of worst linear function.
% Worst linear function in SAS see \url{https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mi_sect027.htm}.
% Note: we're talking about missing data only, not the combined data (that autocorrelation is very high, as is the autocorrelation of deductively imputed values, like in the texp example). 
% "Applications of MICE with lowly correlated data therefore inject a lot of noise into the system. Hence, the auto-correlation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent. Thus, the incorporation of noise into the imputed data has pleasant side-effect of speeding up convergence" \citep{buur18}, par. 4.5.

@@

- $\widehat{R}$ threshold: Replicate simulation study and build a decision rule to solve the problem with $\widehat{R}$.
% "The monitoring statistic was computed for mean monthly earnings at each iteration in chains of length k = 200. Since calculation of the statistic requires M parallel sequences, m = 5 such chains were constructed. This value of m was informed by the preferred choice given in the literature on multiple imputation. The monitoring statistic computed at each iteration is presented in Figures 15 to 17 for each of the three missingness mechanisms. The red vertical line denotes ten iterations" \cite[p.~49]{lace07}.

@@

This research project consists of an investigation into algorithmic convergence of MI algorithms. I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

The aim is to evaluate whether the imputation procedure has converged. The primary research interest is in determining whether GR statistic $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data. We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest.  The simulation diagnostics are as recommended by Stef van Buuren \cite{buur18}. That is, average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations. Also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd}. % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

@@

Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data. We expect that the simulation diagnostics will indicate proper imputations before $\widehat{R}$ will.

%% -- Simulation Approach ---------------------------------------------------------------

\section{Simulation Approach}

In this study, 1000 MCMC simulations are performed. The simulation set-up consists of several steps, see pseudo-code below. Simulation code is available in online appendix XYZ on Github.

\begin{Code}
# pseudo-code of simulation 
simulate data with missingness
for (number of simulation runs from 1 to 1000)
  for (number of iterations from 1 to 100)
    impute the missingness
    compute convergence diagnostic
    perform analysis
    pool results
    compute simulation diagnostics
save convergence and simulation diagnostics
\end{Code}

The simulated data is a finite population of $N=1000$. The variables are dependent variable $Y$, independent variable $X$, and covariates $Z_1$ and $Z_2$. The quantity of scientific interest is the estimated regression coefficient of $X$ on $Y$ in linear regression model $Y \sim X+Z_1+Z_2$. The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$. Outcome variable $Y$ is deduced from the predictor variables as follows:

\begin{align*}
\begin{pmatrix}X\\
Z_{1}\\
Z_{2}\\
\epsilon
\end{pmatrix} &\sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,&
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 10
\end{pmatrix}
\end{bmatrix}\\[2\jot]
Y &=  2 \times X + 0.5 \times Z_1 - 1 \times Z_2 + \epsilon
\end{align*}

After data generation, the complete data is 'amputed'. That is, the \pkg{mice} function \fct{ampute} is used to impose an MCAR missingness mechanism upon the data. The probability to be missing is the same for all cells, namely XYZ\%. Table XYZ shows a summary of the generated complete data, and the amputed data. The amputed data is the starting point of each of the 1000 simulations per simulation condition.

Missing data points are imputed with \pkg{mice} in \proglang{R}. All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains. The number of iterations is varied over simulations ('maxit' argument between 1 and 100). For each number of iterations, simulation and convergence diagnostics are aggregated over 1000 MCMC simulations. 
% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p. 4)

%% -- Results ---------------------------------------------------------------

\section{Results}

Figure \ref{resultssim} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{subsetresults}.

\begin{figure}[h]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{res3.pdf}
  }      
  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
    \label{resultssim}
\end{figure}

From the simulation diagnostics (...) it appears that as little as three iterations is sufficient to draw valid inference. Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations. Apparently, they are not perfect for MI data.

Moreover, $\widehat{R}$ could  theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ). How could $\widehat{R}$ smaller than 1 occur? The number of simulations is smaller than in 'regular' MCMC processes. Therefore, the '$(n-1/n)$' correction factor can influence the estimated potential scale reduction factor. This downwards bias is in the opposite direction than expected: "The mixture-of-sequences variance, V, should stabilize as a function of n. (Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)" \cite[p~438]{broo98}.
